Population structure and phenotypic variation of *Sclerotinia sclerotiorum* from dry bean throughout the United States, and from Mexico, France, and Australia

### 2017-06-28 doc/manuscript/manuscript.Rmd
update general formatting

A population study from Australian canola fields showed evidence of sexual outcrossing of *S. sclerotiorum* isolates [@sexton2004microsatellite].
Another study characterized three populations of *S. sclerotiorum* from lettuce in California (CA) and Ontario (ON), and pea/lentil in Washington (WA) [@malvarez2007new].
While all three populations were  clonal, the index of association (IA) based on DNA fingerprint was closer to zero for CA than it was for the other populations.
Populations from Australian sunflower with evidence of clonal reproduction were reported, and despite populations sampled from fields separated 50 to 400 km and from both head and basal rots, all showed similar gene diversity, with authors concluding isolates representing different geographies belong to one large genetic population of *S. sclerotiorum* [@ekins2011population].

@@

We analyzed *S. sclerotiorum* isolates collected from dry bean cultivars from 2003 to 2012 from different geographic locations in America, Europe and Australia using microsatellite markers to delineate genetic differences and reproduction behavior.
The specific aims of our study were:  a) To assess association of phenotypic data (virulence) and environmental data (bean cultivar, region, field, year) on genotypic variation of *S. sclerotiorum* isolates, b) To test the hypothesis that *S. sclerotiorum* populations of dry beans originated from different continents were genetically different, and c) although previous studies show that *S. sclerotiorum* populations are largely clonal, few compare populations from different continents **(Lehner and Mizubuti, 2017)**, c) To identify localized genotypes representative of each location and also investigate if there are widespread genotypes common among locations. d) To test potential for sexual reproduction in sub populations of the pathogen isolates under different conditions such as year or field in each region, year/region, field/region/year, and e) Test if pathogen within-field populations differentiate over time.

@@

A distance-based redundancy analysis (dbRDA) [@legendre1999distance] was performed with the function capscale in the vegan 2.4-3 package in R (https://CRAN.R-project.org/package=vegan) to delineate relative contribution of phenotypic, spatial, and temporal components in influencing genetic composition of the populations. dbRDA is a method of conducting constrained ordinations on data matrices consisting of non-Euclidean distance measures.
RDA is a multivariate linear regression analysis of independent (explanatory: eg. environmental and phenotypic) variables on dependent (non-explanatory or response) variables.
In our case, a pairwise Bruvo's [@bruvo2004simple] distance matrix was generated from microsatellite data using poppr 2.4.1 [@kamvar2014poppr; @kamvar2015novel] and employed it as the dependent matrix for RDA analysis.
To identify environmental variable(s) correlated with genetic variation, a forward-backward selection process was applied with the vegan function ordistep to identify the set of variables that explained more variation than predicted with the global model containing all variables.
An ANOVA was then performed to test for significance of the reduced model and marginal effects using 999 permutations.
The varpart function of vegan was used to determine variation partitioning of explanatory variables.

@@

To assess the connectivity of shared genotypes across populations, we first tabulated the multilocus genotypes shared between at least two populations (defined as states or countries) with the poppr function mlg.crosspop() [@kamvar2014poppr].
From these data, we were able to construct a graph with populations as nodes and shared genotypes as edges (connections) between the nodes using the R packages igraph version 1.0.1 **(Csardi and Nepusz 2006)**, dplyr version 0.5.0 **(Wickham and Francois 2016)**, and purrr version 0.2.2 **(Wickham 2016)**.
Each node was weighted by the fraction of shared MLGs.
Each edge represents a single MLG, but because a single MLG could cross populations, that MLG may be represented by (n*(n-1))/2 edges where n represents the number of populations it crosses.
Edges were weighted by 1- Psex, where Psex is the probability of encountering the same genotype via two independent meiotic events (Arnaud-Haond 2007, Parks and Werth 1993).
We then used the cluster_optimal() function from igraph to define communities **(Csardi and Nepusz 2006)**.
The graph was plotted using the R packages ggplot2 version 2.2.1 **(Wickham 2009)** and ggraph version 1.0.0 **(Pedersen 2017)**.
To ensure that we were able to capture the same community signal, we additionally ran this analysis including the 5 polymorphic markers described above.

@@

DAPC is a multivariate method designed to identify and describe clusters of genetically related individuals.
Discriminant analysis maximizes separation between groups while minimizing variation within groups [@jombart2010discriminant].
The find.clusters function in Adegenet 2.0.1 [@jombart2008adegenet] uses a k-means clustering algorithm to identify the optimum number of clusters.
It was not able to identify a single optimum and identified a range of optimal clusters between 5-15.
Consequently, DAPC was performed several times with different optimal number of clusters to determine whether isolates (N = 318) fell into each cluster have any biological meaning such as belonging to a particular host or came from the same region etc.  the best summaries for the total isolates (N = 318).
Since dbRDA identified region, host, and sampling year as the top three variables influencing microsatellite data, DAPC was also performed on data defined according to each of these variables, separately.
Isolates (N = 318) were assigned to different levels of the three factors (region, host, year) and assessed predefined group size (original number of isolates at each level of the factor), post group size (number of isolates assigned by the DAPC algorithm) and the proportion of the original isolates assigned to the original group after analysis.
This was performed to determine the suitability of explanatory variables (i.e. region, host, and year) in describing the clusters generated.
Clone-corrected isolates sampled in different years from the white mold nurseries (wmn) were tested for population differentiation over time by DAPC analysis.
Only WA, MI, and NE isolates were qualified since we filtered out the states that had at least $\geq$ 7 samples/year with at least two years of data for the analysis.
Lastly, we pooled all isolates, clone corrected them (N = 165) and performed DAPC using the function find.clusters as described above to delineate how many clusters were formed by these unique genotypes.
AMOVA was used to test the significance of clusters resulted by different DAPC analyses using GenAlEx version 6.41 [@peakall2012genalex].
PhiPT which is an analogue of Wright's FST was used to ascertain the significance of populations.
An FST value of 0.05 or less is generally considered as reasonably low and may be interpreted to mean that structuring between populations is weak [@jost2008gst].
<!-- **(Balloux and Ugon-Moulin, 2002; Wright, 1978)**.   -->

@@

Genotypic evenness [@grunwald2003analysis], which is a measure of distribution of genotype abundances and unbiased gene diversity (Hexp) [@nei1978estimation] were calculated using poppr package in R.
Microsatellite data were also tested for evidence of linkage disequilibrium, where significant disequilibrium between pairs of loci in a population is attributed to linkage between loci and may be an indication of clonal reproduction if loci are physically unlinked, whereas linkage equilibrium can be an indication of outcrossing populations.
The linkage disequilibrium of clone-corrected data was determined by calculating the index of association ($I_{A}$) and its standardized version ($\bar{r}_d$) and a permutation approach with 999 permutations was used to test whether observed values were significantly different from data randomized at each locus [@agapow2001indices; @brown1980multilocus].
Value of IA tends to increase with the number of loci and therefore we made our inferences based on the , which standardizes the number of loci tested and considered a more robust method [@agapow2001indices].
Higher observed  or IA values would indicate clonally reproducing populations.

@@

Community analysis of the MLGs shared between populations revealed 4 communities with a modularity of 0.17 (Fig. \@ref(community-graph)): A coastal community (CA, OR, WA, and NY), a mid-west community (CO, NE, ND, MI), and an international community (MN, France, Australia).
Analysis with 16 loci resulted in the removal of the NY node because it no longer shared a genotype with OR, but the same overall community structure was present with a modularity of 0.2 (Supplementary Figure S1Z).
The international community appears to be driven by MLG 4, which is shared between all three populations and has a Psex value of 2.87 x 10-5, in contrast to the abundant MLG 25, which has a Psex value of 0.0168.

@@

The number of observed alleles per locus ranged anywhere from 2 to 10 (Table 2).
All loci were informative at cut off level of 2/N where N was the total number of individuals.
The genetic diversity indices of *S. sclerotiorum* populations from different states of the USA, France, Mexico and Australia based on SSR analysis are given in Table 5.
The 165 unique MLGs identified among all 366 isolates are given in Supplementary Table 1.
Gene diversity (Hexp) of populations from different regions ranged from 0.259 to 0.596 (Table 5).
Evenness of genotypes of all regions was high, with the exception of ND, where a single genotype was represented by 20 isolates.
There was one private allele in the WA population (204 bp allele of 55-4 locus) and two private alleles in the ND population (two 366 bp alleles in 9-2 locus and two 220 bp alleles in 12-2 locus).

@@

Since find.clusters function of Adegenet library did not result in a clear-cut number of clusters for 318 isolates, we arrived at six (6) as the best number of populations (Fig. 9) for our data after examining results of DAPC for different number of clusters.
Three clusters, namely clusters one, three, and six were formed separately from the rest.
The cluster 3 of the DAPC plot consisted entirely of MX isolates (N =13/15).
Rest of the two MX isolates were grouped in cluster 1.
Cluster 6 had 50 isolates in total representing mainly of NE (N = 15/37), MI (N = 12/58), and CO (N = 12/34).
All CA (N = 18) isolates and 13/58 WA isolates dominated the cluster 1.
AMOVA on these six clusters resulted 44% molecular variance among populations with P = 0.001.
Pairwise comparison showed significant difference (P = 0.001) for all pairs.
The DAPC plot after pooling all isolates together and removing clones (N =165) was compared with the plot of 318 isolates looked similar to Fig. 8C.
The number of clusters was set to 6 for the analysis which resulted in a DAPC having similar topology to plot in Fig. 9 with 318 isolates(results not shown).
Isolates of MX grouped separately while another distinct group had 24 isolates representing CO, NE, and ND.

@@

Interestingly, all Mexican genotypes were unique to that region.
On the other hand CA genotypes were present in the West Coast states of OR and WA which may be due to movement of sclerotia with planting material.
Surprisingly, France and Australian genotypes were also isolated in the USA.
MLG 4 (see supplemental Table 1) was recorded from France, Australia and MN while France and USA isolates had four common MLGs.
Similarly, five MLGs were recorded common to both Australia and USA isolates.
However, there were six unique genotypes in clone-corrected France population (N= 6/14) and three genotypes in CA population (N = 3/15).
MLGs from Australian region, our samples were limited to 6, which is not a sufficient number to make inferences.
Our results indicated that different continents were not a defining factor for microsatellite genotypes of *S. sclerotiorum* isolated from dry bean.
This may be due to involuntary movement of sclerotia by human activities and probably because of recent speciation of *S. sclerotiorum*. 
A phylogenetic study on Sclerotinaceae showed that most of the sclerotial lineages including different species of Sclerotinia had evolved relatively recently from a common ancestor [@carbone1993ribosomal].
Few multilocus genotypes were unique to each sampling state of the USA but most of them were common among several states (Supplemental Table 1).
For instance, Michigan state from which we had the largest collection of isolates (N = 62) consisted of 43 MLGs.
Of these, 32 MLGs were present in at least one of the 5 states of NE, ND, WA, CO, MN; 4 MLGs were present in France isolates; and 11 were unique to MI sites (see Supplemental Table 1 for unique MLGs in different regions).
When selecting *S. sclerotiorum* isolates for screening dry bean germplasm for disease resistance, plant breeders may find above information useful.
A study on genetic diversity of *S. sclerotiorum* from multiple host crops in the North central region of the USA revealed there was little evidence for genetic isolation of this pathogen by host crop or by distance [@aldrich-wolfe2015genetic].
Though our isolates didn't come from different crops, we had isolates from many resistant and susceptible dry bean cultivars, which also didn't show genetic differentiation according to AMOVA.
It seems Mexican pathogen isolates are different from the rest and warrants further investigation using a larger population.
Some France genotypes were common in USA isolates but at the same time 43% (6/14) of the France genotypes were unique to that region indicating white mold resistant cultivars produced in the USA might not work in that region.
Geographic distance between different states of the USA seem not a barrier for gene transfer of *S. sclerotiorum* isolates though unique genotypes did prevail at a low frequency in each state.
The genetic diversity of regions with >10 isolates had similar values (0.43 - 0.59) except for Mexican isolates.
Hexp of MX was 0.26 and indicated less exposure of sampling sites to outside inocula.
Test for linkage disequilibrium showed evidence of clonal reproduction for MX isolates (Ia = 2.61, P.Ia = 0.001,  = 0.37, P. = 0.001).
This may have been another reason for low gene diversity.

>>>>

A population study from Australian canola fields showed evidence of sexual outcrossing of *S. sclerotiorum* isolates [@sexton2004microsatellite].
Another study characterized three populations of *S. sclerotiorum* from lettuce in California (CA) and Ontario (ON), and pea/lentil in Washington (WA) [@malvarez2007new].
While all three populations were  clonal, the index of association ($I_A$) based on DNA fingerprint was closer to zero for CA than it was for the other populations.
Populations from Australian sunflower with evidence of clonal reproduction were reported, and despite populations sampled from fields separated 50 to 400 km and from both head and basal rots, all showed similar gene diversity, with authors concluding isolates representing different geographies belong to one large genetic population of *S. sclerotiorum* [@ekins2011population].

@@

We analyzed *S. sclerotiorum* isolates collected from dry bean cultivars from 2003 to 2012 from different geographic locations in America, Europe and Australia using microsatellite markers to delineate genetic differences and reproduction behavior.
The specific aims of our study were:  a) To assess association of phenotypic data (virulence) and environmental data (bean cultivar, region, field, year) on genotypic variation of *S. sclerotiorum* isolates, b) To test the hypothesis that *S. sclerotiorum* populations of dry beans originated from different continents were genetically different, and c) although previous studies show that *S. sclerotiorum* populations are largely clonal, few compare populations from different continents [@lehner2017sclerotinia], c) To identify localized genotypes representative of each location and also investigate if there are widespread genotypes common among locations. d) To test potential for sexual reproduction in sub populations of the pathogen isolates under different conditions such as year or field in each region, year/region, field/region/year, and e) Test if pathogen within-field populations differentiate over time.

@@

A distance-based redundancy analysis (dbRDA) [@legendre1999distance] was performed with the function `capscale()` in the *vegan* 2.4-3 package in R (https://CRAN.R-project.org/package=vegan) to delineate relative contribution of phenotypic, spatial, and temporal components in influencing genetic composition of the populations. dbRDA is a method of conducting constrained ordinations on data matrices consisting of non-Euclidean distance measures.
RDA is a multivariate linear regression analysis of independent (explanatory: eg. environmental and phenotypic) variables on dependent (non-explanatory or response) variables.
In our case, a pairwise Bruvo's [@bruvo2004simple] distance matrix was generated from microsatellite data using *poppr* 2.4.1 [@kamvar2014poppr; @kamvar2015novel] and employed it as the dependent matrix for RDA analysis.
To identify environmental variable(s) correlated with genetic variation, a forward-backward selection process was applied with the vegan function `ordistep()` to identify the set of variables that explained more variation than predicted with the global model containing all variables.
An ANOVA was then performed to test for significance of the reduced model and marginal effects using 999 permutations.
The `varpart()` function of vegan was used to determine variation partitioning of explanatory variables.

@@

To assess the connectivity of shared genotypes across populations, we first tabulated the multilocus genotypes shared between at least two populations (defined as states or countries) with the *poppr* function `mlg.crosspop()` [@kamvar2014poppr].
From these data, we were able to construct a graph with populations as nodes and shared genotypes as edges (connections) between the nodes using the R packages *igraph* version 1.0.1 **(Csardi and Nepusz 2006)**, *dplyr* version 0.5.0 **(Wickham and Francois 2016)**, and *purrr* version 0.2.2 **(Wickham 2016)**.
Each node was weighted by the fraction of shared MLGs.
Each edge represents a single MLG, but because a single MLG could cross populations, that MLG may be represented by (n\*(n-1))/2 edges where n represents the number of populations it crosses.
Edges were weighted by $1-P_{sex}$, where $P_{sex}$ is the probability of encountering the same genotype via two independent meiotic events [@arnaud2007standardizing, @parks1993study].
We then used the `cluster_optimal()` function from *igraph* to define communities **(Csardi and Nepusz 2006)**.
The graph was plotted using the R packages *ggplot2* version 2.2.1 **(Wickham 2009)** and *ggraph* version 1.0.0 **(Pedersen 2017)**.
To ensure that we were able to capture the same community signal, we additionally ran this analysis including the 5 polymorphic markers described above.

@@

DAPC is a multivariate method designed to identify and describe clusters of genetically related individuals.
Discriminant analysis maximizes separation between groups while minimizing variation within groups [@jombart2010discriminant].
The `find.clusters()` function in *adegenet* 2.0.1 [@jombart2008adegenet] uses a k-means clustering algorithm to identify the optimum number of clusters.
It was not able to identify a single optimum and identified a range of optimal clusters between 5-15.
Consequently, DAPC was performed several times with different optimal number of clusters to determine whether isolates (N = 318) fell into each cluster have any biological meaning such as belonging to a particular host or came from the same region etc.  the best summaries for the total isolates (N = 318).
Since dbRDA identified region, host, and sampling year as the top three variables influencing microsatellite data, DAPC was also performed on data defined according to each of these variables, separately.
Isolates (N = 318) were assigned to different levels of the three factors (region, host, year) and assessed predefined group size (original number of isolates at each level of the factor), post group size (number of isolates assigned by the DAPC algorithm) and the proportion of the original isolates assigned to the original group after analysis.
This was performed to determine the suitability of explanatory variables (i.e. region, host, and year) in describing the clusters generated.
Clone-corrected isolates sampled in different years from the white mold nurseries (wmn) were tested for population differentiation over time by DAPC analysis.
Only WA, MI, and NE isolates were qualified since we filtered out the states that had at least $\geq$ 7 samples/year with at least two years of data for the analysis.
Lastly, we pooled all isolates, clone corrected them (N = 165) and performed DAPC using the function find.clusters as described above to delineate how many clusters were formed by these unique genotypes.
AMOVA was used to test the significance of clusters resulted by different DAPC analyses using GenAlEx version 6.41 [@peakall2012genalex].
PhiPT which is an analogue of Wright's FST was used to ascertain the significance of populations.
An $F_{st}$ value of 0.05 or less is generally considered as reasonably low and may be interpreted to mean that structuring between populations is weak [@jost2008gst].
<!-- **(Balloux and Ugon-Moulin, 2002; Wright, 1978)**.   -->

@@

Genotypic evenness [@grunwald2003analysis], which is a measure of distribution of genotype abundances and unbiased gene diversity ($H_{exp}$) [@nei1978estimation] were calculated using *poppr* package in R [@kamvar2014poppr].
Microsatellite data were also tested for evidence of linkage disequilibrium, where significant disequilibrium between pairs of loci in a population is attributed to linkage between loci and may be an indication of clonal reproduction if loci are physically unlinked, whereas linkage equilibrium can be an indication of outcrossing populations.
The linkage disequilibrium of clone-corrected data was determined by calculating the index of association ($I_{A}$) and its standardized version ($\bar{r}_d$) and a permutation approach with 999 permutations was used to test whether observed values were significantly different from data randomized at each locus [@agapow2001indices; @brown1980multilocus].
Value of $I_A$ tends to increase with the number of loci and therefore we made our inferences based on the , which standardizes the number of loci tested and considered a more robust method [@agapow2001indices].
Higher observed  or $I_A$ values would indicate clonally reproducing populations.

@@

Community analysis of the MLGs shared between populations revealed 4 communities with a modularity of 0.17 (Fig. \@ref(community-graph)): A coastal community (CA, OR, WA, and NY), a mid-west community (CO, NE, ND, MI), and an international community (MN, France, Australia).
Analysis with 16 loci resulted in the removal of the NY node because it no longer shared a genotype with OR, but the same overall community structure was present with a modularity of 0.2 (Supplementary Figure S1Z).
The international community appears to be driven by MLG 4, which is shared between all three populations and has a $P_{sex}$ value of 2.87 x 10^-5^, in contrast to the abundant MLG 25, which has a $P_{sex}$ value of 0.0168.

@@

The number of observed alleles per locus ranged anywhere from 2 to 10 (Table 2).
All loci were informative at cut off level of 2/N where N was the total number of individuals.
The genetic diversity indices of *S. sclerotiorum* populations from different states of the USA, France, Mexico and Australia based on SSR analysis are given in Table 5.
The 165 unique MLGs identified among all 366 isolates are given in Supplementary Table 1.
Gene diversity ($H_{exp}$) of populations from different regions ranged from 0.259 to 0.596 (Table 5).
Evenness of genotypes of all regions was high, with the exception of ND, where a single genotype was represented by 20 isolates.
There was one private allele in the WA population (204 bp allele of 55-4 locus) and two private alleles in the ND population (two 366 bp alleles in 9-2 locus and two 220 bp alleles in 12-2 locus).

@@

Since `find.clusters()` function of *adegenet* did not result in a clear-cut number of clusters for 318 isolates, we arrived at six (6) as the best number of populations (Fig. 9) for our data after examining results of DAPC for different number of clusters.
Three clusters, namely clusters one, three, and six were formed separately from the rest.
The cluster 3 of the DAPC plot consisted entirely of MX isolates (N =13/15).
Rest of the two MX isolates were grouped in cluster 1.
Cluster 6 had 50 isolates in total representing mainly of NE (N = 15/37), MI (N = 12/58), and CO (N = 12/34).
All CA (N = 18) isolates and 13/58 WA isolates dominated the cluster 1.
AMOVA on these six clusters resulted 44% molecular variance among populations with P = 0.001.
Pairwise comparison showed significant difference (P = 0.001) for all pairs.
The DAPC plot after pooling all isolates together and removing clones (N =165) was compared with the plot of 318 isolates looked similar to Fig. 8C.
The number of clusters was set to 6 for the analysis which resulted in a DAPC having similar topology to plot in Fig. 9 with 318 isolates(results not shown).
Isolates of MX grouped separately while another distinct group had 24 isolates representing CO, NE, and ND.

@@

Interestingly, all Mexican genotypes were unique to that region.
On the other hand CA genotypes were present in the West Coast states of OR and WA which may be due to movement of sclerotia with planting material.
Surprisingly, France and Australian genotypes were also isolated in the USA.
MLG 4 (see supplemental Table 1) was recorded from France, Australia and MN while France and USA isolates had four common MLGs.
Similarly, five MLGs were recorded common to both Australia and USA isolates.
However, there were six unique genotypes in clone-corrected France population (N= 6/14) and three genotypes in CA population (N = 3/15).
MLGs from Australian region, our samples were limited to 6, which is not a sufficient number to make inferences.
Our results indicated that different continents were not a defining factor for microsatellite genotypes of *S. sclerotiorum* isolated from dry bean.
This may be due to involuntary movement of sclerotia by human activities and probably because of recent speciation of *S. sclerotiorum*. 
A phylogenetic study on Sclerotinaceae showed that most of the sclerotial lineages including different species of Sclerotinia had evolved relatively recently from a common ancestor [@carbone1993ribosomal].
Few multilocus genotypes were unique to each sampling state of the USA but most of them were common among several states (Supplemental Table 1).
For instance, Michigan state from which we had the largest collection of isolates (N = 62) consisted of 43 MLGs.
Of these, 32 MLGs were present in at least one of the 5 states of NE, ND, WA, CO, MN; 4 MLGs were present in France isolates; and 11 were unique to MI sites (see Supplemental Table 1 for unique MLGs in different regions).
When selecting *S. sclerotiorum* isolates for screening dry bean germplasm for disease resistance, plant breeders may find above information useful.
A study on genetic diversity of *S. sclerotiorum* from multiple host crops in the North central region of the USA revealed there was little evidence for genetic isolation of this pathogen by host crop or by distance [@aldrich-wolfe2015genetic].
Though our isolates didn't come from different crops, we had isolates from many resistant and susceptible dry bean cultivars, which also didn't show genetic differentiation according to AMOVA.
It seems Mexican pathogen isolates are different from the rest and warrants further investigation using a larger population.
Some France genotypes were common in USA isolates but at the same time 43% (6/14) of the France genotypes were unique to that region indicating white mold resistant cultivars produced in the USA might not work in that region.
Geographic distance between different states of the USA seem not a barrier for gene transfer of *S. sclerotiorum* isolates though unique genotypes did prevail at a low frequency in each state.
The genetic diversity of regions with >10 isolates had similar values (0.43 - 0.59) except for Mexican isolates.
$H_{exp}$ of MX was 0.26 and indicated less exposure of sampling sites to outside inocula.
Test for linkage disequilibrium showed evidence of clonal reproduction for MX isolates (Ia = 2.61, P.Ia = 0.001,  = 0.37, P. = 0.001).
This may have been another reason for low gene diversity.
